
\renewcommand{\thechapter}{2}

\chapter{Programmable Property-Based Testing}\label{chap:ppbt}

In the introduction, I introduced a simple API for writing property-based tests. Define
your predicate, write your generator, put them together into a \code{Property} to be
tested by the testing framework. In this chapter, I will walk the reader through an
exposition of the rough edges of the existing APIs for writing properties, followed by
an identification of a core problem in the design of existing PBT libraries. I will
present the ramifications of the problem followed by a solution from our paper with the
same title as the chapter, ``Programmable Property-Based Testing''~\cn.

\section{How to write properties?}

The introduction had a deliberately simple property, single input, no preconditions, no
need for any configuration changes or augmentations. From here on, I will use
Binary-Search Tree properties taken from How to Specify It!~\cite{Hughes2019HowTS} and
have been integrated into ETNA~\cite{ETNA} as case studies. I will also start using
Rocq for the exposition as it is the language I will eventually present the solution
in. Rocq hosts QuickChick~\cite{Pierce:SF4}, one of the most faithful ports of
QuickCheck, keeping much of the original machinery, the types and algorithms of
QuickCheck in place.

\begin{code}[language=Rocq]
Definition insert_correct (t: Tree) (k: nat) (k': nat) (v: nat) : option bool :=
    isBST t
    -=> find k' (insert k v t) = if k =? k' then Some v
                                 else find k' t.
\end{code}

The property \code{InsertPost} states that for a binary-search tree, inserting an
element should, if the keys are equal (which should happen rarely but sometimes), let
us find the newly inserted element (which checks if any immediate insertion can fail);
if the keys aren't equal, it shouldn't change the resulting value we get for any of the
other possible keys. This property fails with \todo{K} different bugs we will go into
detail in the next chapter. For now, we will focus on how to test it rather than what
those tests would reveal.

Let us initially consider how we would write the property to be tested by QuickChick:

\begin{code}[language=Rocq]
Definition prop_insert_correct :=
  forAll arbitrary (fun (t: Tree) =>
  forAll arbitrary (fun (k: nat)  =>
  forAll arbitrary (fun (k': nat) =>
  forAll arbitrary (fun (v: nat)  =>
  prop_InsertPost t k k' v)))).
\end{code}

There's a small caveat to the using \code{Arbitrary} typeclass, which is that each type
can have at most one typeclass instance, so we cannot really specialize for a specific
scenario. We can switch \code{arbitrary} in the property definition with specialized
generators, we can even define the \shrinker{shrinkers} and \printer{printers}
ourselves if we would like using \code{forAllShrinkShow}:

\begin{code}[language=Rocq]
Definition prop_insert_correct :=
  forAllShrinkShow gen shr pri (fun (t: Tree) =>
  forAll (choose -100 100)     (fun (k: nat)  =>
  forAll (choose -100 100)     (fun (k': nat) =>
  forAll (choose -100 100)     (fun (v: nat)  =>
  prop_InsertPost t k k' v)))).
\end{code}

Well, that seems fine, but ideally we would not even generate the \code{k}, \code{k'},
and \code{v} if the tree \code{t} doesn't pass the precondition \code{isBST}, however
right now we generate everything before starting to check. We could solve this by
turning to \code{forAllShrinkShowMaybe} (this is starting to look awfully like
\code{AbstractPropertyFactory} at this point):

\begin{code}[language=Rocq]
Definition prop_insert_correct :=
  forAllShrinkShowMaybe (
    fun _ =>
        <todo>
  ) shr pri (fun (t: Tree) =>
  forAll (choose -100 100)   (fun (k: nat)  =>
  forAll (choose -100 100)   (fun (k': nat) =>
  forAll (choose -100 100)   (fun (v: nat)  =>
    find k' (insert k v t) = if k =? k' then Some v
                             else find k' t)))).
\end{code}

There are a couple more changes we can do, such as adding callbacks to \code{collect}
statistics from the tests, and we can configure some other aspects of the testing by
the configuration options QuickChick gives us, but otherwise, this is all the
flexibility we get from the testing library. The more pressing question is, \emph{is
    this enough?}

Property-Based Testing literature tells us it is not. Looking at
Table~\ref{tab:pbt-frameworks}, we see significant fragmentation within languages;
Haskell hosts QuickCheck, SmallCheck~\cite{RuncimanNL08}, Hedgehog~\cite{Hedgehog} as
well as LeanCheck\cn, and falsify~\cite{falsify} that is not mentioned in the table.
These new frameworks emerge as a response to the limits of flexibility in the existing
ones\cn. These limitations arise at points that surpass the prescience of the library
authors in the requirements of the users, because the dominant design paradigm of the
property languages for PBT libraries fundamentally \emph{fixes} the underlying property
runner, albeit with a wide range of configurations for the user. All possible behaviors
and configuration options must be programmed into the library internals with no ability
to change or extend them in the user space. In the next section, I will introduce a
variety of property runners from the literature that diverges from the original
property runner of QuickCheck illustrated in
Figure~\ref{fig:quickcheck-property-runner-intro}.

\section{Property Runners in Literature}



\input{tables/pbt-frameworks-full}